Topological Equivariant Artist Model for Visualization Library Architecture

### 2025-01-03 paper/main.tex
homemeomorphism

The fiber category $\mathcal{\dfiberc}$ is also equipped with a bifunctor because it is a monoidal category and this functor provides a method for combining fiber types. The bifunctor allows $\otimes$ us to express fields that contain complexly typed values. For example, dates can be represented as three fields $\dfiber_{year} \times \dfiber_{month} \times \dfiber_{day}$ or a composite fiber field $\dfiber_{year} \times \dfiber_{month} \times \dfiber_{day} = \dfiber_{date}$. The $\otimes$ encapsulates both the sets of values associated with each fiber $\{y \in \mathbb{I}\vert 1992 \leq y \leq 2025\} \times \{m \in \mathbb{I}\vert 1 \leq m \leq 12 \} \times \{d \in \mathbb{I}\vert 1 \leq d \leq 31 \}$ and the composition function $ \otimes: \dfiber_{year} \times \dfiber_{month} \times \dfiber_{day} \rightarrow \dfiber_{date}$
could include a constraint to only return dates that have the right number of days for each month. The bifunctor also composes the morphisms associated with each category into a morphism on the composite category $(\dfunct_{year}, \dfunct_{month}, \dfunct_{day}) = \dfunct_{date}$.

@@

\subsection{Homeomorphism}
As mentioned in \autoref{sec:related-work:continuity}, preserving the topology of a visualization means that each discrete piece of differentiable visual information corresponds to a dist element of the dataset\cite{ziemkiewiczEmbeddingInformationVisualization2009} in a way where the organization of elements is preserved. A generalization of this condition is the idea that the graphic space can be collapsed into the data indexing space, which means that the data base space is a deformation retraction of the graphic base space\cite{hatcherAlgebraicTopology2002}. By defining the indexing look up function \vindex, introduced in \autoref{sec:atct:sheaves}, to be
\begin{equation}
  \vindex: \dbase \times I \rightarrow \dbase\;s.t\;\vindex(\dbasepoint) = \dbasepoint\;\forall \dbasepoint \in \gbase
\end{equation}

we can assert that the data space \dbase\ acts as an indexing space into \gbase\ such that knowing the location in one space yields the location on the other and any point in either base space or graphic space has a corresponding point in the other space.
\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{deform_retract.png}
  \caption{The graphic base space $\gbase$ is collapsible to the line $\dbase$ such that every band $(\dbasepoint_i, [0,1])$ on $\gbase$ maps to corresponding point $\dbasepoint_i \in \dbase$. The band $[0,1]$ determines the thickness of a rendered line for a given point $\dbasepoint_{i}$ by specifying how pixels corresponding to that point are colored. \label{fig:construction:xi}}
\end{figure}

For example, as shown in \autoref{fig:construction:xi}, a line is 1D but is a 2D glyph on a screen; therefore the graphic space $\gbase$ is constructed by multiplying the base space $\dbase$ with an interval $[0,1]$. Because $\gbase$ is collapsible into $\dbase$, every band $(\dbasepoint_i, [0,1])$ corresponds to a point in the base space $\dbasepoint_i \in \dbase$. The first coordinate $\alpha=\dbasepoint_i$ provides a lookup to retrieve the associated visual variables. The second coordinate, which is a point in the interval $\beta=[0,1]$. Together they are a point $\gbasepoint=(\alpha,\beta) \in gbase$ in the graphic base space. This point $\gbasepoint$ is the input into the graphic section $\gsection(\gbasepoint)$ that is used to determine which pixels are colored, which in turn determines the thickness, texture, and color of the line.

Specifying the artist as a natural transform (\autoref{df:natural-transform}) on sheaves means that the artist preserves the presheaf and sheaf constraints (\autoref{def:atct:presheaf}, \autoref{def:atct:sheaf}) when mapping data into graphics. The set of possible visualizations for a larger set of values must include all visualizations for a subset of those values  $\vartist(\Gamma(\dbase_1, \dtotal)) \xrightarrow{i^*} \vartist(\Gamma(\dbase_2, \dtotal))$ when more values may exist on the subset of the index $\dbase_2 \xhookrightarrow{i} \dbase_1$. While the curves in \autoref{fig:atct:presheaf} are meant as stand-ins for the data, they also serve to illustrate that the relationship is identical in the graphic space. The artist preserves the \textit{locality} axiom in that identical datasets $\dsection^{a} = \dsection^{b}$ generate identical specifications $\vartist(\dsection^a) = \vartist(\dsection^b)$. The artist respects \textit{gluing} in that graphics are composable, as described in \autoref{sec:artist:operators}, such that a set of artists that generate graphics for parts of a dataset are equivalent to an artist that generates a graphic for the whole dataset. Managing subsets of data and on demand data in a structure preserving way is necessary to support visualizing streaming data \cite{krstajicVisualizationStreamingData2013} and techniques that rely on shifting or resampling the indexing space, such as  panning and zooming \cite{NekrasovskiEvaluationPanZoom2006}.

>>>>

The fiber category $\mathcal{\dfiberc}$ is also equipped with a bifunctor because it is a monoidal category and this functor provides a method for combining fiber types. The bifunctor allows $\otimes$ us to express fields that contain complexly typed values. For example, dates can be represented as three fields $\dfiber_{year} \times \dfiber_{month} \times \dfiber_{day}$ or a composite fiber field $\dfiber_{year} \times \dfiber_{month} \times \dfiber_{day} = \dfiber_{date}$. The $\otimes$ encapsulates both the sets of values associated with each fiber $\{y \in \mathbb{I}\vert 1992 \leq y \leq 2025\} \times \{m \in \mathbb{I}\vert 1 \leq m \leq 12 \} \times \{d \in \mathbb{I}\vert 1 \leq d \leq 31 \}$ and the composition function $ \otimes: \dfiber_{year} \times \dfiber_{month} \times \dfiber_{day} \rightarrow \dfiber_{date}$
would include a constraint to only return dates that have the right number of days for each month. The bifunctor also composes the morphisms associated with each category into a morphism on the composite category $(\dfunct_{year}, \dfunct_{month}, \dfunct_{day}) = \dfunct_{date}$.

@@

\subsection{Homeomorphism}

As introduced in \autoref{sec:related-work:continuity}, the topology of a dataset is preserved in the visualization if there is a homeomorphic map between the graphic and data indexing space. Because multiple graphics can correspond to one data record and every graphic must have a corresponding record \cite{tufteVisualDisplayQuantitative2001,ziemkiewiczEmbeddingInformationVisualization2009}, we define \textcolor{functor}{indexing look up functor} \vindexc\ as a deformation retraction\cite{hatcherAlgebraicTopology2002} of the graphic indexing space \gbase\ into the data indexing space \dbase. This means that every point in \gbase\ can be mapped, non-uniquely, to a point in \dbase\:

\begin{equation}
  \vindexc: \dbasec \times I \rightarrow \dbasec\;s.t\;\vindexc(\gbasepointc) = \dbasepointc\;\forall \dbasepointc \in \gbasec
\end{equation}
The graphic indexing space is defined as the data space multiplied by an identity $\gbase = \dbase\times I$ because it allows us to encode the data index into the graphic index. For example, for a 2D space the graphic basepoint is defined as $\gbasepoint=(\alpha,\beta) \in \dbasepoint_i\times I$ . This means that the first coordinate $\alpha=\dbasepoint_i$ provides a lookup into the data space and the second coordinate is a point on the interval $\beta \in I$ and together they are input into the graphic function $\gsection(\gbasepoint)$ which returns how the pixel at $\gbasepoint$ is set.

\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{xi_zoom.pdf}
  \caption{The function $\xi$ maps each point in each orange band on each graphic indexing space $\gbase_0, \gbase_1, \gbase_2$ to the same point $\dbasepoint_i$ in the circle $\dbase$. As shown in \autoref{fig:atct:morphisms:sheaf}, the parts of the graphic indexed by the orange bands are visualizing parts of the record at $\dbasepoint_i$ \label{fig:artist:xi}}
\end{figure}

In \autoref{fig:artist:xi} each orange band is defined as $\frac{7\pi}{4}\times [0,1]$ such that each point on each band maps to the same point $\vindex|_{\frac{7\pi}{4}\times [0,1]} = \frac{7\pi}{2}$. As illustrated in \autoref{fig:atct:morphisms:sheaf}, each band acts as an indexer into its corresponding image because the section \gsection\ takes as input coordinates $(\frac{7\pi}{2}, \beta), \beta \in [0,1]$ in each orange band and returns the settings for the pixel it maps into. As illustrated in \autoref{fig:artist:xi}, the definition of $\vindex$ allows for some flexibility in the shape of $\gbase$. An annulus $\gbase_0$ is used for the green circle in \autoref{fig:atct:morphisms:sheaf} because that cyclical structure is in the graphic and in \dbase\, but rectangles $\gbase_1, \gbase_2$ are used for the $\sin$ and $\cos$ curves because they are linear. Despite discarding the cyclical information embedded in \dbase\, the spaces $\gbase_1$ and $\gbase_2$ are structure preserving because they satisfy the constraint on \vindex\ that every graphic element maps back into a data element \cite{ziemkiewiczEmbeddingInformationVisualization2009}.

The artist is specified as a natural natural transform (\autoref{def:natural-transform}) on sheaves because it means that the artist can assume that subsets of the data behave in the same manner as the whole when the artist preserves the presheaf and sheaf constraints (\autoref{def:atct:presheaf}, \autoref{def:atct:sheaf}) when mapping data into graphics. Preserving the presheaf constraints means that the set of possible visualizations for a larger set of values must include all visualizations for a subset of those values  $\vartist(\Gamma(\dbase_1, \dtotal)) \xrightarrow{i^*} \vartist(\Gamma(\dbase_2, \dtotal))$ when more values may exist on the subset of the index $\dbase_2 \xhookrightarrow{i} \dbase_1$. While the curves in \autoref{fig:atct:presheaf} are meant as stand-ins for the data, they also serve to illustrate that the relationship is identical in the graphic space. The artist preserves the \textit{locality} axiom in that identical datasets $\dsection^{a} = \dsection^{b}$ generate identical specifications $\vartist(\dsection^a) = \vartist(\dsection^b)$. The artist respects \textit{gluing} in that graphics are composable, as described in \autoref{sec:artist:operators}, such that a set of artists that generate graphics for parts of a dataset are equivalent to an artist that generates a graphic for the whole dataset. Managing subsets of data and on demand data in a structure preserving way is necessary to support visualizing streaming data \cite{krstajicVisualizationStreamingData2013} and techniques that rely on shifting or resampling the indexing space, such as  panning and zooming \cite{NekrasovskiEvaluationPanZoom2006}.
